Testing Charts: viewers' perceptual accuracy in surveys

### 2023-07-20 SDSS-2023/testing-charts-accuracy.qmd
pdf rendered, small changes to address kr's questions

\kr{The mackinlay thing feels like maybe we are overemphasizing the ranking done by cleveland/mcgill and then mackinlay - I think we could emphasize less if we want}

Our work -- as that of Cleveland and McGill and Heer and Bostock -- centers around studying data visualization design choices and their impact on viewer behavior and accuracy of viewers' responses.
We seek to answer whether it is possible to reproduce some of their findings in the context of a survey with a large, nationally-representative set of respondents, and within that context we focus on the following research questions:

@@

We would expect that comparing unaligned tiles is a harder task (with correspondingly lower levels of accuracy) than a comparison of aligned tiles, with the framing given by the context of the other tiles in the same column mitigating some of this difficulty.
@fig-tasks-explain gives an overview of the comparisons of tasks 1 through 3 and the closest corresponding tasks in Cleveland and McGill.
\kr{The updated sketch is great here!! One note -- can we move the label for the "positions along the same, but unaligned scale", it's not aligned with the others}

@@

Task 1 was presented to viewers in one of two rounds (Rounds 1-2), with four color scheme variations each presented to 50% of respondents in each round.
There were no significant differences in respondent accuracy or behavior across color schemes, and we combine them here into one set of stimulus responses to compare structural differences.
\kr{Should we put the color variations and a test for significance in accuracy/a visual in the appendix/supplementary materials? }

@@

The images utilized in the third task again represent the same data as in the first two tasks, and the overall image has dimensions which are identical to the images in the first two tasks, displaying at the same size during the survey.
However, the aspect ratio of the plotting area is adjusted to increase the length in pixels of the stacked bar elements, and decrease the relative thickness \kr{wording?} of each bar. 
\kr{Do we have a measurement of the pixels for this? I think a sentence on that would close out this section nicely.}

@@

\hh{XXX I think what you are trying to say is that we do not know, if participants have seen the stimuli multiple times, but if they did, there was at least a month in between?} \kr{Yes that's what I was trying to say here!!}

Given the nature of pulling a sample from the panel for each Omnibus round, there is a possibility that some participants may have been included in multiple rounds; however, our collected data is not a longitudinal or panel study, so we do not have repeated responses from the same participants across each of our tests.
Each of the tests was run at least a month apart, so if respondents did participate in multiple rounds, there was at least a one month gap between viewing each visual stimulus.

@@

When combining responses that were gathered during distinct rounds of the Omnibus survey, we make a weighting adjustment to ensure that weights in each sample are properly calibrated to the population total across all rounds in the resulting model.

@@

We *combine* (rather than cumulate) surveys $S_1$ and $S_2$, as described in @omuircheartaighCombiningSamplesVs2002, by multiplying weights in $S_1$ and $S_2$ by $\lambda$ and $1 - \lambda$ respectively.
$$
\lambda = \frac{n_1/d_1}{n_1/d_1 + n_2/d_2},
$$ where $n_1$ and $n_2$ are the nominal sample sizes and $d_1$ and $d_2$ are the design effects for the estimators.
Here, $d_1$ and $d_2$ are estimated as $$
d_1 = 1 + CV(w_i \in S_1)^2 \ \ \ \text{ and } \ \ \ d_2 = 1 + CV(w_i \in S_2)^2
$$ where $CV$ is the coefficient of variation of the weights within each sample, and is estimated as in @kishSurveySampling1965 :

@@

Note that @omuircheartaighCombiningSamplesVs2002 estimate $\lambda$ separately for any combination of race/ethnicity by sex.
We employ that strategy whenever we include demographic variables in the analysis, otherwise we use a single adjustment for the weights.

@@

\kr{We need to be careful about not making it seem like the rounds are the same as the tasks, since we have 3 rounds and 3 tasks. We describe it, but I feel like this table kind of muddies the waters. Could we instead do a table of respondents by task?} \hh{does this comment still stand after the re-organization? ... I would suggest to move it up to the survey methods and add the lambdas} \kr{I think moving it here helps, I did also add another table below. Let me know what you think.}

| Name    | Date       |      \# Participants |                effective sample size |         Sum of weights $\sum_i w_i$ | $\lambda_i$ |
|---------------|---------------|--------------:|--------------:|--------------:|--------------:|
| Round 1 | April 2022 | `r nrow(round1_adj)` | `r round(neff(round1_adj$weight),1)` | `r round(sum(round1_adj$weight),1)` | $\lambda_1$  =  `r round(lambdas[1],3)` |
| Round 2 | May 2022   | `r nrow(round2_adj)` | `r round(neff(round2_adj$weight),1)` | `r round(sum(round2_adj$weight),1)` | $\lambda_2$  = `r sprintf("%.3f",lambdas[2])` |
| Round 3 | Jun 2022   | `r nrow(round3_adj)` | `r round(neff(round3_adj$weight),1)` | `r round(sum(round3_adj$weight),1)` | $\lambda_3$  = `r round(lambdas[3],3)` |
| **Total**        |    ---        |    `r nrow(round_123)`   |  `r round(neff(round_123$weight),1)`   |    |  $\lambda_1w_1 + \lambda_2w_2 + \lambda_3w_3$ |

@@

A total of `r sum(c(nrow(round1_adj), nrow(round2_adj), nrow(round3_adj)))` respondents participated across the three rounds. The number of responses and corresponding effective sample sizes in each round are shown in \autoref{tbl-rounds}. All responses were combined into one set of survey responses, with adjusted combined sample weights and indicators for which task each respondent was exposed to. The resulting sample sizes corresponding to each task are shown in \autoref{tbl-tasks}. \kr{Is this table useful? I want to be clear about the difference between tasks and rounds and make sure it's clear what our resulting sizes are..}

@@

Distributions of demographic characteristics across each survey round are shown in \autoref{fig-rounds-demographics}. \kr{Split by rounds or by tasks? I feel like since we later split by task, maybe we should show the demos by task? Or we can just talk about how Round 3 was split randomly along demos for Tasks 2/3?}

@@

round_123 <- round_123 %>% group_by(round) %>% 
  mutate(
    weight_pct_round = weight/sum(weight)*100
  )

gender <- round_123 %>% 
  ggplot(aes(x = gender, weight = weight_pct_round)) + geom_bar() + 
  ylab("Percent") + 
  ggtitle("Gender") +
  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  xlab("") +
  facet_grid(.~round)

age <- round_123 %>% 
  ggplot(aes(x = age4, weight = weight_pct_round)) + geom_bar() + 
  ylab("Percent") + 
  ggtitle("Age ") +
  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  xlab("") + ylim(c(0,35)) +
  facet_grid(.~round)

education <- round_123 %>% 
  ggplot(aes(x = educ5, weight = weight_pct_round)) + geom_bar() + 
  ylab("Percent") + 
  ggtitle("Education levels") +
  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  xlab("") + ylim(c(0,35)) +
  facet_grid(.~round)

income <- round_123 %>% 
  ggplot(aes(x = income4, weight = weight_pct_round)) + geom_bar() + 
  ylab("Percent") + 
  ggtitle("Income levels") +
  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  xlab("") + ylim(c(0,35)) +
  facet_grid(.~round)

@@

@@

```{r}
#| label: fig-demographics
#| fig-cap: "Panelist's demographics matter, particularly, when the task difficulty increases. For aligned tiles, gender, age, and education are not significant factors. However, income levels do have - a small - effect. When income levels increase, the percentage of wrong answers (orange) decreases, while the percentage of correct answers (blue) increases slightly (significant at below 0.05). The more difficult task of comparing unaligned tiles, demographics are more significant.  XXX With increasing levels in education, we see a small drop in correct answers, as well as a small drop in wrong answers, for a significant increase in 'they are the same'."
#| fig-subcap: 
#|   - Comparisons of aligned tiles
#|   - Comparisons of unaligned tiles
#| fig-height: 4
#| fig-width: 7
#| layout-nrow: 2
#| out-width: 100%

@@

```{=tex}
\hh{@tbl-demographics doesn't survive the formatting to pdf very well - need to shorten the row names or put into multiple lines} \hh{table rotates in pdf}
```

>>>>

\kr{The mackinlay thing feels like maybe we are overemphasizing the ranking done by cleveland/mcgill and then mackinlay - I think we could emphasize less if we want}
\hh{my point was to emphasize that Mackinlay is a foundational paper (rightfully so), but not based on any data beyond Cleveland}

<!--Our work -- as that of Cleveland and McGill and Heer and Bostock -- -->

Our work also centers on studying data visualization design choices and their impact on viewer behavior and accuracy of viewers' responses.
We seek to answer whether it is possible to reproduce some of their findings in the context of a survey with a large, nationally-representative set of respondents, and within that context we focus on the following research questions:

@@

We would expect that comparing unaligned tiles is a harder task (with correspondingly lower levels of accuracy) than a comparison of aligned tiles, with the framing given by the context of the other tiles in the same column mitigating some of this difficulty.
@fig-tasks-explain gives an overview of the comparisons of tasks 1 through 3 and the closest corresponding tasks in Cleveland and McGill.
\kr{The updated sketch is great here!! One note -- can we move the label for the "positions along the same, but unaligned scale", it's not aligned with the others}
\hh{that was on purpose :(  that goes beyond Cleveland and McGill ... I'll see what I can do about that}

@@

Task 1 was presented to viewers in one of two rounds (Rounds 1-2), with four color scheme variations each presented to 50% of respondents in each round. 
There were no significant differences in respondent accuracy or behavior across color schemes, and we combine them here into one set of stimulus responses to compare structural differences.
\kr{Should we put the color variations and a test for significance in accuracy/a visual in the appendix/supplementary materials? } \hh{we could do that, but I'm not sure that we need to mention color at all.}

@@

The images utilized in the third task again represent the same data as in the first two tasks, and the overall image has dimensions which are identical to the images in the first two tasks, displaying at the same size during the survey.
\hh{The length of bars is increased to fit the new dimensions. The height of the tiles are adjusted accordingly (from 50 to 30 pixels) to keep the overall area the same as before}
<!--However, the aspect ratio of the plotting area is adjusted to increase the length in pixels of the stacked bar elements, and  decrease the relative thickness \kr{wording?} of each bar. -->
\kr{Do we have a measurement of the pixels for this? I think a sentence on that would close out this section nicely.}\hh{like that?}

@@

<!--\hh{XXX I think what you are trying to say is that we do not know, if participants have seen the stimuli multiple times, but if they did, there was at least a month in between?} \kr{Yes that's what I was trying to say here!!}-->

\hh{The Omnibus sample is not longitudinal in nature, i.e.\ we do not have any information about whether the same panelists were included in multiple rounds. While there is a (small) chance that panelists are included in multiple rounds of our data, there was at least a one month gap between viewing each visual stimulus, and for the purposes of the analysis we will consider data from these viewings as independent.  }

<!-- Given the nature of pulling a sample from the panel for each Omnibus round, there is a possibility that some participants may have been included in multiple rounds; however, our collected data is not a longitudinal or panel study, so we do not have repeated responses from the same participants across each of our tests.
Each of the tests was run at least a month apart, so if respondents did participate in multiple rounds, there was at least a one month gap between viewing each visual stimulus.
-->

@@

When combining responses from different rounds of the Omnibus survey, we calibrate weights in each round to the total population  across all rounds in the resulting model.

@@

We *combine* (rather than cumulate) two surveys $S_1$ and $S_2$, as described in @omuircheartaighCombiningSamplesVs2002, by multiplying weights in $S_1$ and $S_2$ by $\lambda$ and $1 - \lambda$ respectively. $\lambda \in [0,1]$ is given as
$$
\lambda = \frac{n_1/d_1}{n_1/d_1 + n_2/d_2},
$$ where $n_1$ and $n_2$ are the nominal sample sizes and $d_1$ and $d_2$ are the design effects for the estimators.
Here, $d_1$ and $d_2$ are estimated as $$
d_1 = 1 + CV(w_i \in S_1)^2 \ \ \ \text{ and } \ \ \ d_2 = 1 + CV(w_i \in S_2)^2,
$$ where $CV$ is the coefficient of variation of the weights within each sample, and is estimated as in @kishSurveySampling1965 :

@@

<!--Note that @omuircheartaighCombiningSamplesVs2002 estimate $\lambda$ separately for any combination of race/ethnicity by sex.
We employ that strategy whenever we include demographic variables in the analysis, otherwise we use a single adjustment for the weights. -->

@@

\kr{We need to be careful about not making it seem like the rounds are the same as the tasks, since we have 3 rounds and 3 tasks. We describe it, but I feel like this table kind of muddies the waters. Could we instead do a table of respondents by task?} \hh{does this comment still stand after the re-organization? ... I would suggest to move it up to the survey methods and add the lambdas} \kr{I think moving it here helps, I did also add another table below. Let me know what you think.}\hh{I like it!}

| Name    | Date       |      \# Participants |                effective sample size |         Sum of weights $\sum_i w_i$ | $\lambda_i$ |
|---------------|---------------|--------------:|--------------:|--------------:|--------------:|
| Round 1 | April 2022 | `r nrow(round1_adj)` | `r round(neff(round1_adj$weight),1)` | `r round(sum(round1_adj$weight),1)` | $\lambda_1$  =  `r round(lambdas[1],3)` |
| Round 2 | May 2022   | `r nrow(round2_adj)` | `r round(neff(round2_adj$weight),1)` | `r round(sum(round2_adj$weight),1)` | $\lambda_2$  = `r sprintf("%.3f",lambdas[2])` |
| Round 3 | Jun 2022   | `r nrow(round3_adj)` | `r round(neff(round3_adj$weight),1)` | `r round(sum(round3_adj$weight),1)` | $\lambda_3$  = `r round(lambdas[3],3)` |
| **Total**        |    ---        |    `r nrow(round_123)`   |  `r round(neff(round_123$weight),1)`   |    |   |

@@

A total of `r sum(c(nrow(round1_adj), nrow(round2_adj), nrow(round3_adj)))` respondents participated across the three rounds. The number of responses and corresponding effective sample sizes in each round are shown in \autoref{tbl-rounds}. All responses were combined into one set of survey responses, with adjusted combined sample weights and indicators for which task each respondent was exposed to. The resulting sample sizes corresponding to each task are shown in \autoref{tbl-tasks}. \kr{Is this table useful? I want to be clear about the difference between tasks and rounds and make sure it's clear what our resulting sizes are..}\hh{yes!}

@@

Distributions of demographic characteristics across each survey round are shown in \autoref{fig-rounds-demographics}. \kr{Split by rounds or by tasks? I feel like since we later split by task, maybe we should show the demos by task? Or we can just talk about how Round 3 was split randomly along demos for Tasks 2/3?}\hh{changed the figure to show demos by task -- do we need confidence intervals? -- or a comparison against census data?}

@@

round_123 <- round_123 %>% group_by(chart) %>% 
  mutate(
    weight_pct_round = weight/sum(weight)*100
  )

gender <- round_123 %>% 
  ggplot(aes(x = gender, weight = weight_pct_round)) + geom_bar() + 
  ylab("Percent") + 
  ggtitle("Gender") +
  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  xlab("") +
  facet_grid(.~chart)

age <- round_123 %>% 
  ggplot(aes(x = age4, weight = weight_pct_round)) + geom_bar() + 
  ylab("Percent") + 
  ggtitle("Age ") +
  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  xlab("") + ylim(c(0,35)) +
  facet_grid(.~chart)

education <- round_123 %>% 
  ggplot(aes(x = educ5, weight = weight_pct_round)) + geom_bar() + 
  ylab("Percent") + 
  ggtitle("Education levels") +
  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  xlab("") + ylim(c(0,35)) +
  facet_grid(.~chart)

income <- round_123 %>% 
  ggplot(aes(x = income4, weight = weight_pct_round)) + geom_bar() + 
  ylab("Percent") + 
  ggtitle("Income levels") +
  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  xlab("") + ylim(c(0,35)) +
  facet_grid(.~chart)

@@

\hh{we could include certainty - I'll look into it}

@@

\hh{XXX the formatting isn't right yet here ... latex problem with large floats ... will fix it.}

```{r}
#| label: fig-demographics
#| fig-cap: "Panelists' demographics matter, particularly, when the task difficulty increases. For aligned tiles, gender, age, and education are not significant factors. However, income levels do have a (small) effect. When income levels increase, the percentage of wrong answers (orange) decreases, while the percentage of correct answers (blue) increases slightly (significant at below 0.05). For the  more difficult task of comparing unaligned tiles, demographics are more significant. Higher levels of education and higher levels of income are associated with a significant increase of panelists choosing a response of 'they are the same', resulting in significant decreases of both correct answers and wrong answers."
#| fig-subcap: 
#|   - Comparisons of aligned tiles
#|   - Comparisons of unaligned tiles
#| fig-height: 4
#| fig-width: 7
#| layout-nrow: 2
#| out-width: 100%

@@
